---
title: "RNA-seq Differential Expression Analysis: Dexamethasone Response in Airway Smooth Muscle"
author: "Arjun Babu S B"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 7, fig.height = 5)
library(DESeq2)
library(airway)
library(ggplot2)
library(pheatmap)
library(gprofiler2)
```

## Introduction

This report analyzes RNA-seq data from the `airway` dataset (Himes et al.,
2014, *PLoS ONE*): human airway smooth muscle cell lines, some treated with
dexamethasone (a glucocorticoid) and some untreated. The goal is to identify
genes whose expression changes significantly with treatment, and to interpret
those changes biologically.

```{r load-data}
data(airway)
colData(airway)
```

## Building the DESeq2 Dataset

Raw counts are combined with sample metadata into a `DESeqDataSet`. The design
formula `~ cell + dex` tells DESeq2 to account for baseline differences
between the four cell lines, so that only the effect of treatment (`dex`) is
tested. Genes with very low counts across all samples are filtered out before
analysis, as they carry little information and only add statistical noise.

```{r build-dds}
dds <- DESeqDataSetFromMatrix(
  countData = assay(airway),
  colData   = colData(airway),
  design    = ~ cell + dex
)

keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

dds <- estimateSizeFactors(dds)
sizeFactors(dds)
```

## Quality Control

A variance-stabilizing transformation puts genes on a comparable scale before
exploratory analysis. PCA and a sample-distance heatmap both check whether
samples separate by treatment, independent of the statistical test to come.

```{r qc-transform}
vsd <- vst(dds, blind = TRUE)
```

```{r pca-plot}
plotPCA(vsd, intgroup = "dex")
```

```{r sample-heatmap}
sampleDists      <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix)
```

Both the PCA and the sample-distance heatmap show samples clustering cleanly
by treatment group, indicating that dexamethasone treatment is a dominant
source of variation in this dataset.

## Differential Expression Analysis

The reference level is set to `"untrt"` so that results describe the effect
of adding dexamethasone, and the full DESeq2 statistical workflow is run.

```{r deseq-test}
dds$dex <- relevel(dds$dex, ref = "untrt")
dds     <- DESeq(dds)
res     <- results(dds)
summary(res)
```

## Volcano Plot

Genes are classified as significantly up- or down-regulated using an adjusted
p-value cutoff of 0.05 combined with a minimum 2-fold change
(`|log2FoldChange| > 1`), to focus on changes that are both statistically
robust and biologically substantial.

```{r rank-results}
res <- res[order(res$padj), ]

res_df      <- as.data.frame(res)
res_df$gene <- rownames(res_df)

res_df$sig <- "Not significant"
res_df$sig[res_df$padj < 0.05 & res_df$log2FoldChange >  1] <- "Up"
res_df$sig[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"
```

```{r volcano-plot}
ggplot(res_df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue",
                                 "Not significant" = "grey")) +
  theme_minimal()
```

## Gene Annotation

Ensembl gene IDs are mapped to readable gene symbols using annotation
metadata already bundled with the `airway` package.

```{r annotate}
gene_symbols <- as.data.frame(rowData(airway))[, c("gene_id", "gene_name")]

res_annotated <- merge(res_df, gene_symbols, by.x = "gene", by.y = "gene_id")
res_annotated <- res_annotated[order(res_annotated$padj), ]

head(res_annotated[, c("gene", "gene_name", "log2FoldChange", "padj")])
```

As a validation check, `CRISPLD2` — the gene highlighted in the title of the
original publication — is confirmed as significant and up-regulated in this
independently reproduced analysis:

```{r crispld2-check}
res_annotated[res_annotated$gene_name == "CRISPLD2", ]
```

```{r save-results}
write.csv(res_annotated, "results_DE_annotated.csv", row.names = FALSE)
```

## Heatmap of Top Differentially Expressed Genes

```{r top-genes-heatmap}
top_hits <- head(res_annotated, 20)
mat      <- assay(vsd)[top_hits$gene, ]
rownames(mat) <- top_hits$gene_name

pheatmap(mat,
         annotation_col = as.data.frame(colData(vsd))["dex"],
         scale = "row")
```

## Pathway Enrichment

Significant genes are tested for enrichment in Gene Ontology (Biological
Process) and KEGG pathway categories using `gprofiler2`, which queries gene
symbols directly against a public enrichment database.

```{r enrichment}
sig_genes <- res_annotated[res_annotated$sig %in% c("Up", "Down"), ]
gene_list <- sig_genes$gene_name

gostres <- gost(query = gene_list, organism = "hsapiens",
                sources = c("GO:BP", "KEGG"))
```

```{r enrichment-results}
go_bp <- gostres$result[gostres$result$source == "GO:BP", ]
go_bp_specific <- go_bp[go_bp$term_size < 500, ]
go_bp_specific <- go_bp_specific[order(go_bp_specific$p_value), ]
head(go_bp_specific[, c("term_name", "p_value", "term_size", "intersection_size")])

kegg_hits <- gostres$result[gostres$result$source == "KEGG", ]
kegg_hits <- kegg_hits[order(kegg_hits$p_value), ]
head(kegg_hits[, c("term_name", "p_value", "term_size", "intersection_size")])
```

## Conclusions

- Dexamethasone treatment produces a clear, strong transcriptional response:
  samples separate cleanly by treatment in both PCA and hierarchical
  clustering, well before any statistical testing.
- Differential expression testing identifies thousands of significantly
  affected genes, including `CRISPLD2`, independently reproducing the
  headline finding of the original publication.
- Pathway enrichment highlights extracellular matrix organization, muscle
  system processes, and inflammatory signaling — biologically consistent
  with known glucocorticoid mechanisms in airway smooth muscle.

**Dataset:** Himes BE, et al. (2014) RNA-Seq Transcriptome Profiling
Identifies CRISPLD2 as a Glucocorticoid Responsive Gene that Modulates
Cytokine Function in Airway Smooth Muscle Cells. *PLoS ONE* 9(6):e99625.
